```{r echo=FALSE}
library(knitr)
options(scipen=999)
opts_chunk$set(comment=NA, fig.width=12, fig.height=6)
```
# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r echo=TRUE}
# Load the data with their associated column type
# download and read the data, convert columns for convenience
read_data <- function() {
    fname = "repdata-data-activity.zip"
    source_url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    if(!file.exists(fname)) {
        download.file(source_url, destfile=fname, method="curl")
    }
    con <- unz(fname, "activity.csv")
    tbl <- read.csv(con, header=T, colClasses=c("numeric", "Date", "factor"))
}
data <- read_data()
str(data)
summary(data)
```

## What is mean total number of steps taken per day?
```{r echo=TRUE}
# Clean NA rows
dataClean <- data[!is.na(data$steps), ]
str(dataClean)
summary(dataClean)

# Returns data aggregated by date
stepsByDay <- function(stepData) {
    tData <- aggregate(stepData$steps, by=list(stepData$date), sum, na.rm=TRUE)
    colnames(tData) <- c("date", "steps")
    tData
}

tData <- stepsByDay(dataClean)
head(tData)

meanStepsByDay <- mean(tData$steps)
medianStepsByDay <- median(tData$steps)
```
The **mean** number of steps taken per day are; `r meanStepsByDay`  
The **median** number of steps taken per day are; `r medianStepsByDay`

In a graph:

```{r echo=TRUE}

plotData <- function(stepData) {
    par(mfrow=c(1,2))
    hist(stepData$steps, xlab="Steps", main="Steps per day Histogram", breaks=10)
    barplot.default(stepData$steps, names.arg=stepData$date, main="Steps per day", ylab="Steps", xlab="Day" )
    abline(h=meanStepsByDay)
    abline(h=medianStepsByDay)    
}

plotData(tData)
```

## What is the average daily activity pattern?
```{r echo=TRUE}
tData <- aggregate(list("Steps Mean"=dataClean$steps), by=list(interval=dataClean$interval), mean)

# Convert to numeric and sort it for plotting
tData$interval <- as.integer(levels(tData$interval))
tData <- tData[order(tData$interval),]

max <- tData[which.max(tData$Steps.Mean), ]

plot(tData$interval, tData$Steps.Mean, type="l", xlab="Interval", ylab="Mean Step", main="Mean steps by interval", col="blue")
abline(v=max[1, "interval"], col="red")
legend('topright', c("Steps", "Maximun"), lty=c(1,1), col=c("blue","red"))
```
The maximun activity happens in interval **`r max[1, "interval"]`** with **`r max[1, "Steps.Mean"]` steps**

## Imputing missing values
```{r echo=TRUE}
# Vector with indexes for rows with NA. NA are only in 'steps'
na_idx <- which(is.na(data$steps))
# Calculation of replacement values to substitute NA values.
na_replacements <- unlist(lapply(na_idx, FUN = function(idx) {
    interval = data[idx, ]$interval
    tData[tData$interval == interval, ]$Steps.Mean
}))


tData <- data
# NA substitution
tData$steps[na_idx] <- na_replacements

summary(tData)

tData <- stepsByDay(tData)
plotData(tData)

meanStepsByDayRep <- mean(tData$steps)
medianStepsByDayRep <- median(tData$steps)
```

The **mean** number of steps taken per day are; `r meanStepsByDayRep`  
The **median** number of steps taken per day are; `r medianStepsByDayRep`

In a graph:

```{r echo=TRUE}
plotData(tData)
```
```{r echo=TRUE}
tData <- data.frame("Original"=c(meanStepsByDay, medianStepsByDay), "Replaced"=c(meanStepsByDayRep, medianStepsByDayRep))
tData$Original <- format(tData$Original, , nsmall = 3)
tData$Replaced <- format(tData$Replaced, , nsmall = 3)
rownames(tData) <- c("mean", "median")
tData

```
So, updating NA values with the interval median steps keep main unchanged while increasing the median to be equal to the mean.

## Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE}
tData2 <- dataClean
# Add a column for the weekday
tData2$weekday <- as.factor(weekdays(tData2$date))

weekend <- subset(tData2, weekday %in% c("Saturday", "Sunday"))
weekday <- subset(tData2, !weekday %in% c("Saturday", "Sunday"))

weekend_spi <- aggregate(list("Steps Mean"=weekend$steps), by=list(interval=weekend$interval), mean, na.rm=TRUE)
weekday_spi <- aggregate(list("Steps Mean"=weekday$steps), by=list(interval=weekday$interval), mean, na.rm=TRUE)

weekend_spi$interval <- as.integer(levels(weekend_spi$interval))
weekday_spi$interval <- as.integer(levels(weekday_spi$interval))

weekend_spi <- weekend_spi[order(weekend_spi$interval),]
weekday_spi <- weekday_spi[order(weekday_spi$interval),]

par(mfrow=c(1,2))
plot(weekend_spi$interval, weekend_spi$Steps.Mean, type="l", xlab="Interval", ylab="Mean Step", main="Weekend", col="blue")
plot(weekday_spi$interval, weekday_spi$Steps.Mean, type="l", xlab="Interval", ylab="Mean Step", main="Weekday", col="blue")

```
As we can see in the figures, the activity is greater in weekends going on through most of the day intervals.